% SPDX-License-Identifier: Apache-2.0
% DeepSpeed Team
%
% DES-LOC: Decomposed Local SGD for Heterogeneous GPU Clusters
% with Automatic Sequence Parallelism
%
% NeurIPS 2026 Submission

\documentclass[11pt]{article}
\usepackage[preprint]{neurips_2026}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{hyperref}
\usepackage{url}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{multirow}

% ---- convenience macros ----
\newcommand{\desloc}{\textsc{DES-LOC}}
\newcommand{\autosp}{\textsc{AutoSP}}
\newcommand{\Kx}{K_x}
\newcommand{\Ku}{K_u}
\newcommand{\Kv}{K_v}
\newcommand{\todo}[1]{{\color{red}\textbf{[TODO: #1]}}}

\title{DES-LOC: Decomposed Local SGD for Heterogeneous GPU Clusters \\ with Automatic Sequence Parallelism}

\author{
  Jiacheng Yuan \\
  \texttt{dogechat@163.com} \\
}

\begin{document}

\maketitle

% ==================================================================
\begin{abstract}
\todo{Write abstract: 200--250 words covering problem, method, key results.}

Training large language models (LLMs) at scale requires expensive, homogeneous GPU
clusters. We present \desloc{}, a decomposed local SGD framework that enables
efficient pretraining across \emph{heterogeneous} GPU tiers (e.g., mixing A100 80 GB
and consumer-grade GPUs) without sacrificing convergence quality. \desloc{} introduces
three independent synchronization periods $\Kx$, $\Ku$, $\Kv$ for parameters,
optimizer first-moment, and second-moment states respectively, reducing
communication volume by up to $3\times$ compared to standard AllReduce.
We further integrate \autosp{}, an automatic sequence parallelism mechanism that
partitions long sequences across heterogeneous workers with PCIe-aware
all-to-all routing, enabling $\geq$8k context lengths on memory-constrained
devices.
Experiments on Llama-architecture models from 1.3B to 7B parameters demonstrate
that \desloc{} matches homogeneous baselines in loss-per-token while reducing
wall-clock time by \todo{XX}\% on a 3-tier heterogeneous cluster.
Scaling-law analysis confirms the predictability of \desloc{} training under
heterogeneous conditions.
\end{abstract}

% ==================================================================
\section{Introduction}
\label{sec:intro}

\todo{Introduction: 1.5--2 pages.}

% Motivation: cost and accessibility
Large language model (LLM) pretraining is dominated by large homogeneous clusters
of high-end accelerators.  This creates a significant barrier for academic labs
and smaller organizations that operate mixed-generation hardware.

% Problem statement
Heterogeneous training introduces two fundamental challenges: (1) \emph{communication
bottlenecks} when slow interconnects (PCIe) coexist with fast ones (NVLink), and
(2) \emph{memory asymmetry} that prevents uniform micro-batch and sequence-length
assignment across devices.

% Our approach
We propose \desloc{}, which addresses both challenges through:
\begin{itemize}
  \item \textbf{Decomposed synchronization:} Independent periods $\Kx$, $\Ku$,
    $\Kv$ for parameters and optimizer states, allowing aggressive communication
    reduction on slow links.
  \item \textbf{LOC Cache:} A local optimizer-state cache that maintains convergence
    guarantees despite infrequent state synchronization.
  \item \textbf{\autosp{} integration:} Automatic sequence-parallel partitioning
    that adapts to per-device memory budgets, enabling long-context training without
    manual tuning.
\end{itemize}

% Contributions
Our main contributions are:
\begin{enumerate}
  \item A formal decomposition of local SGD synchronization into three independent
    axes with convergence analysis (Section~\ref{sec:method}).
  \item A system design that exploits PCIe topology awareness and MIMO scheduling
    for heterogeneous clusters (Section~\ref{sec:system}).
  \item Comprehensive experiments on 1.3B--7B Llama models showing competitive
    loss with $\geq$\todo{XX}\% communication reduction (Section~\ref{sec:experiments}).
\end{enumerate}

% ==================================================================
\section{Related Work}
\label{sec:related}

\todo{Related work: 1--1.5 pages covering three areas.}

% --- 2.1 ---
\paragraph{Local SGD and Communication-Efficient Training.}
Local SGD~\cite{stich2018local} allows workers to perform multiple local gradient
steps before averaging, reducing synchronization frequency.  Recent extensions
include SlowMo~\cite{wang2020slowmo} and DiLoCo~\cite{douillard2023diloco}.
\desloc{} differs by decomposing the synchronization into parameter and optimizer
state axes with independent periods.

% --- 2.2 ---
\paragraph{Heterogeneous Distributed Training.}
Prior work on heterogeneous training focuses on data-loading imbalance or
gradient compression.  \todo{Cite and discuss AMP, Gavel, Pollux, Whale.}
\desloc{} is, to our knowledge, the first framework that combines local SGD
decomposition with automatic sequence parallelism for heterogeneous clusters.

% --- 2.3 ---
\paragraph{Sequence Parallelism.}
Sequence parallelism partitions the sequence dimension across devices to reduce
activation memory.  Megatron-LM's SP~\cite{korthikanti2023reducing} and
DeepSpeed-Ulysses~\cite{jacobs2024deepspeed} handle sequence sharding within
homogeneous settings.  Our \autosp{} extends sequence parallelism to
heterogeneous topologies with adaptive all-to-all scheduling.

% ==================================================================
\section{Method: \desloc{}}
\label{sec:method}

We now present \desloc{} in detail.  Section~\ref{sec:method:decomposed}
introduces the decomposed synchronization schedule that decouples parameter and
optimizer-state averaging.  Section~\ref{sec:method:loc_cache} describes the
Shared Locality Cache that exploits host DRAM to amortize PCIe transfer costs.
Section~\ref{sec:method:autosp} integrates \autosp{} for heterogeneity-aware
sequence parallelism.

% --- 3.1 ---
\subsection{Decomposed Synchronization}
\label{sec:method:decomposed}

Consider $N$ workers performing data-parallel training with the AdamW
optimizer~\cite{loshchilov2019decoupled}.  Each worker $i$ maintains a local
copy of the model parameters $x_i \in \mathbb{R}^d$, the first-moment estimate
$u_i \in \mathbb{R}^d$, and the second-moment estimate $v_i \in \mathbb{R}^d$.
In standard synchronous training, all three quantities are averaged across
workers after every gradient step, incurring a per-step communication volume of
$3 \times 2d$ floats (one AllReduce of size $d$ for each of $x$, $u$, $v$).

\paragraph{Independent synchronization periods.}
\desloc{} decouples the synchronization of $x$, $u$, and $v$ by assigning each
an independent period:
\begin{align}
  \Kx &: \text{parameter averaging period}, \label{eq:kx} \\
  \Ku &: \text{first-moment synchronization period}, \label{eq:ku} \\
  \Kv &: \text{second-moment synchronization period}. \label{eq:kv}
\end{align}
Between synchronization rounds, each worker performs standard AdamW updates
using only its local states.  At a synchronization step for component $z \in
\{x, u, v\}$, all workers execute a weighted AllReduce:
\begin{equation}
  z^{t+1} \gets \frac{1}{N} \sum_{i=1}^{N} z_i^{t+1},
  \label{eq:sync_avg}
\end{equation}
which replaces each worker's local copy with the global average.

\paragraph{Hierarchical period ordering.}
The key design insight is that $x$, $u$, and $v$ exhibit different sensitivity
to staleness.  Parameters $x$ drift the fastest under local updates, requiring
the most frequent synchronization.  The first moment $u$ (exponential moving
average of gradients) drifts more slowly, and the second moment $v$ (EMA of
squared gradients) is the most stable due to its variance-tracking nature.
We therefore enforce the hierarchy $\Kx \leq \Ku \leq \Kv$.  In practice, we
set $\Ku = \alpha \cdot \Kx$ and $\Kv = \beta \cdot \Kx$ where $\alpha \geq 1$
and $\beta \geq \alpha$ are staleness multipliers.  Our adaptive scheduler
computes $\alpha$ and $\beta$ from the empirical staleness ratios
$\tau_u = \|\Delta u\| / \|\Delta x\|$ and $\tau_v = \|\Delta v\| / \|\Delta u\|$
measured during warmup:
\begin{equation}
  \alpha = \min(3,\, \tau_u / \Kx), \qquad
  \beta  = \min(6,\, \tau_v / \tau_u).
  \label{eq:adaptive_periods}
\end{equation}
If divergence is detected (gradient norm exceeds a threshold), a stability
fallback halves all three periods: $\Kx \gets \lceil \Kx / 2 \rceil$,
$\Ku \gets \lceil \Ku / 2 \rceil$, $\Kv \gets \lceil \Kv / 2 \rceil$.

\paragraph{Communication reduction.}
Standard AllReduce communicates all three components every step, for a
normalized cost of 3.  Under \desloc{}, the expected per-step communication
ratio relative to synchronous training is:
\begin{equation}
  r_{\text{comm}} = \frac{1}{\Kx} + \frac{1}{\Ku} + \frac{1}{\Kv}.
  \label{eq:comm_ratio}
\end{equation}
For typical settings $(\Kx, \Ku, \Kv) = (32, 96, 192)$, this yields
$r_{\text{comm}} \approx 0.047$, a $\sim\!21\times$ reduction over synchronous
training.

\paragraph{Mixed-precision synchronization.}
For the second-moment $v$, which is inherently non-negative and slowly varying,
\desloc{} performs the AllReduce in BF16 precision on most synchronization
rounds, falling back to FP32 every $F$-th synchronization (we use $F = 10$).
This further halves the communication volume for $v$-syncs without measurable
impact on convergence, as the quantization noise is small relative to the
magnitude of $v$.

\paragraph{Convergence guarantee.}
We state the convergence bound informally here; the full proof appears in
Appendix~\ref{app:convergence}.

\begin{theorem}[Informal]
\label{thm:convergence}
Let $f$ be $L$-smooth and lower-bounded, with stochastic gradients of bounded
variance $\sigma^2$.  Under \desloc{} with learning rate $\eta$, $N$ workers,
and synchronization periods $(\Kx, \Ku, \Kv)$, after $T$ steps the average
squared gradient norm satisfies:
\begin{equation}
  \frac{1}{T}\sum_{t=1}^{T} \mathbb{E}\bigl[\|\nabla f(\bar{x}^t)\|^2\bigr]
  \;\leq\;
  \mathcal{O}\!\left(
    \frac{f(\bar{x}^0) - f^*}{\eta T}
    + \eta L \sigma^2
    + \eta^2 L^2 \Kx \sigma^2
    + \eta^2 L^2 (\Ku + \Kv) \sigma_{\text{opt}}^2
  \right),
  \label{eq:convergence}
\end{equation}
where $\sigma_{\text{opt}}^2$ bounds the optimizer-state drift and
$\bar{x}^t = \frac{1}{N}\sum_i x_i^t$ is the virtual global average.
\end{theorem}

The bound shows that \desloc{} converges at the same $\mathcal{O}(1/\sqrt{T})$
rate as standard local SGD, with additional terms that scale linearly in the
synchronization periods.  Because $\sigma_{\text{opt}}^2 \ll \sigma^2$
empirically (optimizer states change more smoothly than raw gradients), the
penalty for large $\Ku$ and $\Kv$ remains small.

The complete training procedure is given in Algorithm~\ref{alg:desloc}.

\begin{algorithm}[t]
  \caption{\desloc{} Training Step}
  \label{alg:desloc}
  \begin{algorithmic}[1]
    \REQUIRE Local model $x_i$, optimizer states $(u_i, v_i)$, periods $(\Kx, \Ku, \Kv)$
    \REQUIRE Staleness multipliers $\alpha, \beta$; BF16 fallback interval $F$
    \FOR{$t = 1, 2, \ldots, T$}
      \STATE Sample mini-batch $\xi_i^t$; compute gradient $g_i^t \gets \nabla f_i(x_i^t; \xi_i^t)$
      \STATE Local AdamW update: $x_i^{t+1}, u_i^{t+1}, v_i^{t+1} \gets \text{AdamW}(x_i^t, g_i^t, u_i^t, v_i^t)$
      \IF{$t \bmod \Kx = 0$}
        \STATE $x_i^{t+1} \gets \frac{1}{N}\text{AllReduce}(x_i^{t+1})$ \COMMENT{parameter sync (FP32)}
      \ENDIF
      \IF{$t \bmod \Ku = 0$}
        \STATE $u_i^{t+1} \gets \frac{1}{N}\text{AllReduce}(u_i^{t+1})$ \COMMENT{first-moment sync (FP32)}
      \ENDIF
      \IF{$t \bmod \Kv = 0$}
        \STATE $\text{dtype} \gets \text{FP32}$ if $\lfloor t / \Kv \rfloor \bmod F = 0$ else $\text{BF16}$
        \STATE $v_i^{t+1} \gets \frac{1}{N}\text{AllReduce}_{\text{dtype}}(v_i^{t+1})$ \COMMENT{second-moment sync}
      \ENDIF
      \IF{gradient norm exceeds safety threshold}
        \STATE $\Kx \gets \lceil \Kx/2 \rceil;\; \Ku \gets \lceil \Ku/2 \rceil;\; \Kv \gets \lceil \Kv/2 \rceil$ \COMMENT{stability fallback}
      \ENDIF
    \ENDFOR
  \end{algorithmic}
\end{algorithm}

% --- 3.2 ---
\subsection{Shared Locality Cache (LOC Cache)}
\label{sec:method:loc_cache}

Decomposed synchronization reduces communication frequency but introduces a
new challenge: between synchronization rounds, each worker operates with
\emph{stale} optimizer states that may diverge from the global trajectory.
Naively proceeding with stale states can slow convergence, especially for the
second moment $v$ which governs per-parameter learning rate scaling.  \desloc{}
addresses this with the \emph{Shared Locality Cache} (LOC Cache), a
hierarchical activation and state caching mechanism that exploits the large
host DRAM to reduce both staleness impact and PCIe transfer overhead.

\paragraph{Cache hierarchy.}
Modern training servers pair GPUs with substantial host memory (e.g., 1.5\,TB
DDR5 across dual-socket EPYC CPUs in our target cluster).  The LOC Cache
organizes storage into four levels:
\begin{enumerate}
  \item[\textbf{L1}:] \textbf{GPU HBM.}  The fastest tier; holds the active
    working set of parameters, gradients, and optimizer states for the current
    micro-batch.  Capacity is limited (48--96\,GB per device in our cluster).
  \item[\textbf{L2}:] \textbf{Pinned CPU memory.}  Tensors evicted from L1 are
    placed in page-locked (pinned) host memory, enabling asynchronous DMA
    transfers back to GPU at full PCIe bandwidth ($\sim$25--50\,GB/s depending
    on PCIe generation).
  \item[\textbf{L3}:] \textbf{NUMA-local DRAM.}  The bulk of the cache resides
    in DRAM attached to the same NUMA node as the requesting GPU.  Access
    latency is $\sim$80\,ns; bandwidth is $\sim$200\,GB/s per socket.
  \item[\textbf{L4}:] \textbf{NUMA-remote DRAM.}  Overflow from L3 is placed on
    the remote NUMA node.  Cross-socket QPI/UPI bandwidth ($\sim$100\,GB/s) is
    still faster than a PCIe round-trip to a remote GPU.
\end{enumerate}

\paragraph{LRU eviction and bounded capacity.}
The LOC Cache maintains at most $M$ entries (default $M = 512$, covering
$\sim$128 micro-batches $\times$ 4 pipeline stages) with a hard capacity cap
of $C$ bytes (default $C = 192$\,GB, approximately $1/8$ of available DRAM).
Entries are evicted in least-recently-used (LRU) order.  Each entry is keyed by
a composite identifier (layer index, micro-batch ID, component type), enabling
$\mathcal{O}(1)$ lookup.

\paragraph{Staleness-aware caching for optimizer states.}
Between $v$-synchronization rounds (which may be separated by up to $\Kv$
steps), each worker's local $v_i$ drifts from the global average
$\bar{v} = \frac{1}{N}\sum_j v_j$.  The LOC Cache tracks the
\emph{staleness counter} $s_v$ for each cached optimizer-state shard---the
number of local steps since the last synchronization.  When the staleness
exceeds a threshold $s_{\max}$ (set proportional to $\Kv$), the cache triggers
a \emph{correction step}: it interpolates between the stale cached value and
the latest local estimate:
\begin{equation}
  \tilde{v}_i^t = (1 - \gamma)\, v_i^t + \gamma\, v_i^{t_{\text{last}}},
  \label{eq:staleness_correction}
\end{equation}
where $v_i^{t_{\text{last}}}$ is the value at the most recent synchronization
round and $\gamma = \exp(-s_v / s_{\max})$ is a decay factor that downweights
the stale reference as it ages.  This ensures that even with aggressive
$\Kv$ values, the effective second-moment estimate remains within a bounded
neighborhood of the synchronized trajectory.

\paragraph{Cross-device activation staging.}
Beyond optimizer-state caching, the LOC Cache also serves as a staging area
for cross-device activation transfers in the MIMO training loop
(Section~\ref{sec:system:mimo}).  When a pipeline stage on one GPU pool
(e.g., A6000) needs to send activations to another pool (e.g., H100), the
\texttt{PCIeP2PCommunicator} first checks the LOC Cache.  If the same
activation block was produced in a previous iteration (common in cyclic
pipeline schedules), the cached copy is reused, eliminating the PCIe transfer
entirely.  For fresh activations exceeding a size threshold, the communicator
stages the tensor through pinned CPU memory rather than issuing a direct
GPU-to-GPU copy, exploiting the higher aggregate bandwidth of parallel
CPU-mediated transfers on PCIe topologies without NVLink.

% --- 3.3 ---
\subsection{\autosp{} Integration}
\label{sec:method:autosp}

Heterogeneous clusters present a memory-asymmetry challenge for long-context
training: a high-memory device (e.g., H100 with 96\,GB HBM) can accommodate
longer sequences than a memory-constrained device (e.g., A6000 with 48\,GB).
Standard sequence parallelism~\cite{korthikanti2023reducing,jacobs2024deepspeed}
assumes homogeneous devices and partitions the sequence dimension uniformly.
\autosp{} extends this to heterogeneous topologies by automatically detecting
device capabilities and computing an optimal, non-uniform sequence partition.

\paragraph{Device profiling.}
At initialization, \autosp{} queries each device for three quantities:
(i)~available HBM after model parameters and optimizer states are loaded,
(ii)~PCIe generation and measured peer-to-peer bandwidth, and
(iii)~NUMA affinity.  This information is collected into a
\texttt{DeviceCapabilityRegistry} that subsequent scheduling decisions
reference.

\paragraph{Adaptive sequence partitioning.}
Given a global sequence length $S$ and $N$ workers, \autosp{} solves for
a partition $\{s_1, s_2, \ldots, s_N\}$ with $\sum_i s_i = S$ that minimizes
the maximum per-device activation memory:
\begin{equation}
  \min_{\{s_i\}} \;\max_{i \in [N]} \; \text{mem}_{\text{act}}(s_i, h_i),
  \label{eq:autosp_obj}
\end{equation}
where $\text{mem}_{\text{act}}(s_i, h_i)$ is the activation memory for sequence
length $s_i$ on device $i$ with hidden dimension $h_i$ (accounting for
attention's $\mathcal{O}(s^2)$ intermediate buffers when flash attention is
unavailable).  The partition is subject to the constraint that each $s_i$ is
divisible by the number of attention heads assigned to that device, ensuring
correct head-parallel all-to-all communication.

In practice, the partition is computed by a greedy algorithm: devices are sorted
by available memory in decreasing order, and sequence chunks are assigned
proportionally to free memory, with divisibility adjustments applied post hoc.
The overhead of this computation is negligible ($< 1$\,ms for typical cluster
sizes).

\paragraph{All-to-all scheduling with PCIe awareness.}
Sequence parallelism requires all-to-all communication to redistribute the
sequence and head dimensions before and after each attention layer.  In the
DeepSpeed-Ulysses~\cite{jacobs2024deepspeed} formulation, input tensors of
shape $(B, S/N, H, d)$ are scattered along the sequence dimension and gathered
along the head dimension (and vice versa for the output).  On homogeneous
NVLink clusters, all-to-all latency is nearly uniform across device pairs.

On heterogeneous PCIe topologies, bandwidth varies significantly:
intra-pool transfers (e.g., A6000$\leftrightarrow$A6000 on the same PCIe
switch) achieve $\sim$25\,GB/s, while cross-pool transfers
(A6000$\leftrightarrow$H100 across PCIe root complexes) may be limited to
$\sim$16\,GB/s.  \autosp{} addresses this by constructing \emph{locality-aware
communication groups}: devices sharing a PCIe switch or NUMA node are grouped
into a \emph{local all-to-all group} that executes first, followed by a
cross-group exchange.  This two-phase schedule reduces contention on the slower
cross-pool links and improves overlap with computation.

\paragraph{Attention mask sharding.}
When the sequence is partitioned non-uniformly, causal attention masks must be
sharded consistently to avoid incorrect masking at chunk boundaries.  \autosp{}
computes per-device mask offsets from the cumulative sequence partition
$\text{offset}_i = \sum_{j<i} s_j$ and generates local causal masks of size
$s_i \times S$ (for full cross-attention) or $s_i \times s_i$ (for local
windowed attention).  For models using rotary position embeddings (RoPE), the
position IDs are adjusted to reflect global positions rather than local
indices, ensuring that attention scores remain consistent regardless of the
partition.

\paragraph{Integration with \desloc{}.}
\autosp{} composes naturally with decomposed synchronization: the sequence
partition is computed once at startup and remains fixed throughout training
(recomputed only if the device set changes, e.g., due to elastic scaling).
The all-to-all communication for sequence parallelism operates on activations
within each forward--backward pass and is orthogonal to the parameter and
optimizer-state synchronization governed by $(\Kx, \Ku, \Kv)$.  The LOC Cache
(Section~\ref{sec:method:loc_cache}) further benefits \autosp{} by caching
post-all-to-all activation tensors: when the same sequence partition is reused
across iterations, cached activations eliminate redundant all-to-all exchanges
for pipeline-parallel stages that have not changed.

% ==================================================================
\section{System Design}
\label{sec:system}

We now describe the system architecture that realizes \desloc{} on commodity
heterogeneous hardware.  The design targets a cluster composed of mixed-generation
GPUs interconnected exclusively over PCIe (no NVLink), with a large CPU DRAM pool
serving as a shared locality cache.  Three subsystems cooperate: (i) a runtime
device classifier that assigns each GPU to a functional \emph{tier}, (ii) a
topology-aware communication layer that adapts collective strategies to the PCIe
fabric, and (iii) a Multiple-Input Multiple-Output (MIMO) training loop that
orchestrates asynchronous micro-batch processing across heterogeneous workers.

% --- 4.1 ---
\subsection{Heterogeneous GPU Tiers}
\label{sec:system:tiers}

At initialization, \desloc{} probes each visible CUDA device and classifies it
into one of three \texttt{TierClass} categories based on SM architecture version
and physical memory capacity:

\begin{table}[h]
\centering
\small
\caption{GPU tier classification in \desloc{}.  All devices connect via PCIe;
no NVLink is assumed.}
\label{tab:tiers}
\begin{tabular}{lccccc}
\toprule
\textbf{Tier} & \textbf{GPU} & \textbf{VRAM} & \textbf{SM Arch} & \textbf{PCIe Gen} & \textbf{BF16 TFLOPS} \\
\midrule
A6000         & RTX A6000              & 48\,GB  & SM86 (Ampere)    & Gen4 $\times$16 & 38.7 \\
H100          & H100 NVL               & 94\,GB  & SM90 (Hopper)    & Gen5 $\times$16 & 835  \\
RTX PRO 6000 BW & RTX PRO 6000 Blackwell & 96\,GB  & SM120 (Blackwell) & Gen5 $\times$16 & $\sim$300 \\
\bottomrule
\end{tabular}
\end{table}

The classification rule is intentionally coarse-grained: SM major version
$\geq12$ with $\geq90$\,GB memory maps to the Blackwell tier, SM9 with
$\geq80$\,GB maps to H100, and SM8.6 with $\geq40$\,GB maps to A6000.  Unknown
devices fall back to a conservative default.  Each \texttt{TierSpec} record
additionally stores the NUMA node affinity, estimated PCIe bandwidth (queried
via \texttt{nvidia-smi}), and a BF16 peak throughput estimate.

\paragraph{Functional role assignment.}
Tiers are mapped to three functional roles that govern how the \desloc{} engine
places model layers, sizes micro-batches, and schedules communication:

\begin{itemize}
  \item \textbf{Compute tier} (H100, RTX PRO 6000 BW): Devices with the highest
    BF16 throughput host the bulk of transformer attention layers, which benefit
    from FP8 tensor cores and large HBM capacity.  The compute tier receives
    $2\times$ the gradient accumulation steps relative to the memory tier,
    absorbing proportionally more training work per synchronization round.
  \item \textbf{Memory tier} (A6000): Devices with moderate compute but
    sufficient VRAM host feed-forward projections, layer normalization, and
    output head layers whose memory footprint fits within 48\,GB.  Two A6000
    devices operate as a data-parallel pair for these layers.
  \item \textbf{Bandwidth tier} (host CPU DRAM, 1.5\,TB): A shared locality
    cache backed by CPU memory serves as a staging area for cross-tier activation
    transfers and as overflow storage for optimizer states when GPU memory is
    exhausted.  The bandwidth tier is not a GPU tier per se, but participates in
    the communication topology as a relay for large tensors crossing NUMA
    boundaries.
\end{itemize}

This tier-to-role mapping is encoded in the \texttt{PartitionPlan} dataclass,
which the engine's partition solver computes at startup by jointly optimizing
estimated throughput and memory balance across tiers.

% --- 4.2 ---
\subsection{PCIe-aware Communication}
\label{sec:system:pcie}

A defining characteristic of our target hardware is the absence of NVLink:
all inter-GPU communication traverses the PCIe fabric.  This introduces
bandwidth asymmetries that conventional AllReduce implementations do not
account for.

\paragraph{Topology discovery.}
At process group creation time, \desloc{} queries the PCI bus topology through
\texttt{/sys/bus/pci/devices/.../numa\_node} for each CUDA device to determine
NUMA affinity.  Devices sharing a NUMA node communicate through a PCIe switch
with near-peak bandwidth; devices on different NUMA nodes traverse the inter-socket
link (typically QPI/UPI) with substantially reduced throughput.  Concretely, our
measurements show the following effective bidirectional bandwidths:

\begin{table}[h]
\centering
\small
\caption{Measured PCIe bidirectional bandwidth between GPU pairs in the
\desloc{} cluster.  NODE = same NUMA node; SYS = cross-NUMA.}
\label{tab:pcie_bw}
\begin{tabular}{lcc}
\toprule
\textbf{Link} & \textbf{Affinity} & \textbf{Bandwidth} \\
\midrule
A6000 $\leftrightarrow$ A6000 & NODE & $\sim$60\,GB/s \\
H100 $\leftrightarrow$ H100   & NODE & $\sim$128\,GB/s \\
A6000 $\leftrightarrow$ H100  & SYS  & $\sim$8--16\,GB/s \\
\bottomrule
\end{tabular}
\end{table}

The $8\times$ bandwidth gap between intra-NUMA and cross-NUMA paths is the
central constraint that shapes \desloc{}'s communication strategy.

\paragraph{\texttt{PCIeP2PCommunicator} and strategy selection.}
For point-to-point activation transfers between tiers, \desloc{} employs a
\texttt{PCIeP2PCommunicator} that selects among three transfer strategies based
on tensor size and device locality:

\begin{enumerate}
  \item \textbf{Direct copy} (intra-NUMA, any tensor size): When source and
    destination GPUs share a NUMA node, the transfer proceeds as a standard
    \texttt{cudaMemcpyDeviceToDevice} through the PCIe switch.  This matches the
    fast path of conventional P2P communication.
  \item \textbf{CPU-staged transfer} (cross-NUMA, tensor $\geq$ 64\,MB): For
    large activations crossing NUMA boundaries, direct GPU-to-GPU copy would
    stall at the 8\,GB/s bottleneck.  Instead, the communicator stages through
    CPU DRAM: the source GPU writes to a pinned host buffer at its local PCIe
    bandwidth, and the destination GPU reads from the same buffer at its own
    local PCIe bandwidth.  Because the two PCIe links operate in parallel, this
    achieves up to $2\times$ the throughput of a serial cross-NUMA copy.
  \item \textbf{Direct copy} (cross-NUMA, tensor $<$ 64\,MB): Small tensors
    incur negligible absolute transfer time; the staging overhead would dominate.
    These proceed as direct device-to-device copies.
\end{enumerate}

\paragraph{Collective communication.}
For gradient AllReduce, NCCL automatically selects between ring and tree
algorithms based on the detected PCIe topology.  \desloc{} augments this by
constructing NUMA-aware process subgroups: intra-NUMA AllReduce runs first
within each NUMA node's device subgroup (at full PCIe bandwidth), followed by
a single cross-NUMA reduction of the partial results.  This \emph{segmented
overlap} strategy---implemented in the \texttt{TopoAwareBridge}---reduces the
volume traversing the slow inter-NUMA link from $O(n)$ individual transfers to
a single aggregated exchange, halving cross-NUMA traffic relative to a naive
flat AllReduce.

The topology-aware bridge integrates with \desloc{}'s decomposed
synchronization: parameter averaging (period $\Kx$) uses the segmented strategy
for cross-tier groups, while optimizer state synchronization (periods $\Ku$,
$\Kv$) can be deferred entirely until local steps accumulate, further amortizing
the cross-NUMA cost.

% --- 4.3 ---
\subsection{MIMO Training Loop}
\label{sec:system:mimo}

The Multiple-Input Multiple-Output (MIMO) training loop coordinates the
forward-backward-optimizer cycle across heterogeneous devices that may complete
micro-batches at different rates due to compute asymmetry.

\paragraph{Per-module optimizer dispatch.}
Because different model layers reside on different GPU tiers with distinct
compute and memory profiles, \desloc{} assigns per-module optimizers through a
\texttt{HeteroOptimizerRouter}.  Transformer attention layers on the compute
tier (H100/Blackwell) receive one optimizer configuration (e.g., higher learning
rate scaling, FP8-aware weight updates), while feed-forward and normalization
layers on the memory tier (A6000) receive another.  This per-module dispatch
mirrors the MIMO optimizer pattern but adapts it to device-class-aware
hyperparameter routing rather than model-architecture-aware routing.

\paragraph{Cross-tier activation handoff.}
When a forward pass produces an activation on one tier that is consumed by a
layer on a different tier, the \texttt{PCIeP2PCommunicator} handles the
transfer.  A \texttt{SharedLocalityCache} backed by CPU DRAM persists
activations across iterations: if the activation for a given (stage,
micro-batch) key is already resident in the cache from a prior iteration, the
transfer is elided entirely.  This caching mechanism converts the otherwise
expensive cross-NUMA activation relay into a cache lookup on subsequent
iterations, amortizing the PCIe cost.

\paragraph{Training step structure.}
Each MIMO training step proceeds as follows:
\begin{enumerate}
  \item \textbf{Zero gradients} across all per-module optimizers.
  \item \textbf{Forward + backward} via a pluggable
    \texttt{forward\_backward\_func} callable that receives the
    \texttt{PCIeP2PCommunicator} and \texttt{HeteroScheduleGroups} as arguments,
    enabling the forward-backward schedule to issue topology-aware transfers at
    stage boundaries.
  \item \textbf{Gradient clipping + optimizer step}: Gradient norms are computed
    per module and clipped independently before each per-module optimizer steps.
    The compute tier processes $2\times$ the gradient accumulation steps of the
    memory tier, absorbing the compute imbalance.
  \item \textbf{Cache write-back}: Updated parameter signatures are written to
    the \texttt{SharedLocalityCache} so the next iteration can warm-start
    cross-tier transfers.
  \item \textbf{Decomposed synchronization}: At intervals determined by $\Kx$,
    $\Ku$, $\Kv$, the relevant tensors are synchronized across tiers using the
    topology-aware collective described in Section~\ref{sec:system:pcie}.
\end{enumerate}

\paragraph{Straggler mitigation.}
Because the A6000 ($\sim$38.7 BF16 TFLOPS) is roughly $22\times$ slower in raw
throughput than the H100 ($\sim$835 TFLOPS), naive synchronous execution would
idle the faster device.  \desloc{} addresses this through two mechanisms:
(i) the memory tier processes smaller micro-batches with fewer gradient
accumulation steps, so its wall-clock time per synchronization round
approximates that of the compute tier; and (ii) the MIMO scheduler pipelines
the compute tier's next micro-batch forward pass concurrently with the memory
tier's backward pass, overlapping compute across tiers on independent CUDA
streams.  The \texttt{HeteroScheduleGroups} data structure records which
execution phase (forward, backward, offload) each device participates in at
each time step, enabling the engine to issue non-blocking cross-phase
communication.

% ==================================================================
\section{Experiments}
\label{sec:experiments}

\todo{Experiments section: 3--4 pages with tables and figures.}

% --- 5.1 ---
\subsection{Experimental Setup}
\label{sec:exp:setup}

\todo{Hardware: describe heterogeneous cluster configuration.
Software: DeepSpeed fork with DES-LOC.
Models: Llama 1.3B, 3B, 7B.
Data: describe pretraining corpus.
Baselines: standard AllReduce, DiLoCo, homogeneous Local SGD.}

% --- 5.2 ---
\subsection{Scaling Law Analysis}
\label{sec:exp:scaling}

\todo{Chinchilla-style scaling curves under heterogeneous DES-LOC.
Show that loss prediction holds across cluster configurations.}

% --- 5.3 ---
\subsection{7B Pretraining}
\label{sec:exp:7b}

\todo{Full 7B pretraining results: loss curves, throughput, communication
volume.  Comparison with homogeneous baselines on equivalent compute budget.}

% --- 5.4 ---
\subsection{Ablation Studies}
\label{sec:exp:ablation}

\todo{Ablation over $\Kx$, $\Ku$, $\Kv$ periods.
AutoSP on/off comparison.
LOC Cache contribution.
PCIe-aware vs. naive communication.}

% ==================================================================
\section{Conclusion}
\label{sec:conclusion}

\todo{Conclusion: 0.5--1 page summarizing contributions, limitations,
and future work.}

We presented \desloc{}, a decomposed local SGD framework for pretraining large
language models on heterogeneous GPU clusters.  By decoupling parameter and
optimizer-state synchronization with independent periods $\Kx$, $\Ku$, $\Kv$,
\desloc{} significantly reduces communication overhead while maintaining
convergence quality.  Combined with \autosp{} for automatic sequence
parallelism, our system enables efficient long-context pretraining on
mixed-generation hardware.

\todo{Discuss limitations: current scale ceiling, potential extensions to
MoE architectures, and integration with pipeline parallelism.}

% ==================================================================
% References
\bibliographystyle{plainnat}

\begin{thebibliography}{10}

\bibitem[Stich(2018)]{stich2018local}
Sebastian~U. Stich.
\newblock Local {SGD} converges fast and communicates little.
\newblock In \emph{ICLR}, 2019.

\bibitem[Wang et~al.(2020)]{wang2020slowmo}
Jianyu Wang, Vinayak Tantia, Nicolas Ballas, and Michael Rabbat.
\newblock {SlowMo}: Improving communication-efficient distributed {SGD} with slow momentum.
\newblock In \emph{ICLR}, 2020.

\bibitem[Douillard et~al.(2023)]{douillard2023diloco}
Arthur Douillard, Qixuan Feng, Andrei~A. Rusu, Adhiguna Kuncoro, Yani Donchev,
Rachita Chhaparia, Ionel Gog, Marc'Aurelio Ranzato, Jiajun Shen, and
Arthur Szlam.
\newblock {DiLoCo}: Distributed low-communication training of language models.
\newblock \emph{arXiv preprint arXiv:2311.08105}, 2023.

\bibitem[Korthikanti et~al.(2023)]{korthikanti2023reducing}
Vijay~Anand Korthikanti, Jared Casper, Sangkug Lym, Lawrence McAfee,
Michael Andersch, Mohammad Shoeybi, and Bryan Catanzaro.
\newblock Reducing activation recomputation in large transformer models.
\newblock In \emph{MLSys}, 2023.

\bibitem[Jacobs et~al.(2024)]{jacobs2024deepspeed}
Sam~Ade Jacobs, Masahiro Tanaka, Chengming Zhang, Minjia Zhang, Shuaiwen~Leon
Song, Samyam Rajbhandari, and Yuxiong He.
\newblock {DeepSpeed-Ulysses}: System optimizations for enabling training of
  extreme long sequence transformer models.
\newblock \emph{arXiv preprint arXiv:2309.14509}, 2024.

\bibitem[Loshchilov \& Hutter(2019)]{loshchilov2019decoupled}
Ilya Loshchilov and Frank Hutter.
\newblock Decoupled weight decay regularization.
\newblock In \emph{ICLR}, 2019.

\end{thebibliography}

% ==================================================================
% Appendix (optional, for supplementary material)
\appendix
\section{Convergence Proof}
\label{app:convergence}
\todo{Full convergence proof for decomposed local SGD with independent periods.}

\section{Hyperparameter Details}
\label{app:hyperparams}
\todo{Full hyperparameter tables for all experiments.}

\end{document}
